---
title: "assignment"
author: "Rebecca Waag"
date: "26 5 2023"
output: html_document
---

# Week12: Chromatin conformation, domains and looping

## Exercise12: Working with distal regulatory elements

```{r}
#load packages
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(AnnotationHub)
  library(ensembldb)
  library(InteractionSet)
  library(epiwraps)
  library(GenomeInfoDb)
})
```

### Download data

Download JUND peaks from human ChIP experiment and predicted distal targets from Salviato et al. 2021

```{r}
#set working directory
setwd("C:/Users/rebec/Desktop/w12")

#download JUND peaks
#download.file("https://www.encodeproject.org/files/ENCFF130QNO/@@download/ENCFF130QNO.bed.gz",dest="JUND.bed.gz")
#import peaks
JUND<-rtracklayer::import("JUND.bed.gz", format="narrowPeak")
#head(JUND)
#length(JUND) #22681 peaks

#download predicted distal targets from Salviato et al. (2021)
#download.file("https://ethz-ins.org/content/hg38.SalviatoDistalEnhancerTargets.GR.rds", "interactions.GR.rds", mode="wb")
interactions <- readRDS("interactions.GR.rds")
#head(interactions)
#length(interactions) #1085167 distal targets
```

```{r}
#load human genome
ah <- AnnotationHub()
ensdb <- ah[["AH95744"]] # human ensembl 104 annotation
```

### Isolate peaks of different distances from the nearest TSS

Isolate the peaks that are nearer (2.5 kb - 10 kb) and those that are farer (>10 kb) from the TSS. 

```{r warning=FALSE}
#use annotateRegions function to isolate the peaks with different distances from TSS

#isolate peaks hat are within 2.5 kb - 10 kb from the nearest TSS
peaks_near <- epiwraps::annotateRegions(JUND, ensdb, proximal = c(2500,10000))
#peaks_near$class
#peaks_near
levels(peaks_near$class) #peaks from "proximal >2500&<=10000bp" are relevant
table(peaks_near$class=="proximal >2500&<=10000bp") #5909 peaks are between 2.5 and 10kb from TSS
peaks_nearer <- peaks_near[which(peaks_near$class=="proximal >2500&<=10000bp")]
#peaks_nearer #GRanges with 5909 ranges

#isolate peaks hat are further than 10 kb from the nearest TSS
peaks_far <- epiwraps::annotateRegions(JUND, ensdb, proximal = c(10000,1e+60))
#peaks_far
#peaks_far$class
levels(peaks_far$class) #peaks from "proximal >10000&<=1e+60bp" are relevant
table(peaks_far$class=="proximal >10000&<=1e+60bp") #10257 peaks are between >10kb frm TSS
peaks_farer <- peaks_far[which(peaks_far$class=="proximal >10000&<=1e+60bp")]
#peaks_farer #GRanges with 10257 ranges

```

5909 peaks are located within 2.5 kb - 10 kb from the TSS, while 10257 peaks are further than 10 kb from the nearest TSS. 

### Characterize peaks

#### Calculate proportion of peaks within 2.5 to 10 kb from the TSS whose predicted target is its nearest TSS

```{r}
# find overlaps between the nearer peaks and distal regulatory elements
overlap_near <- findOverlaps(peaks_nearer, interactions)
#head(overlap_near)
#overlap_near
length(overlap_near) #3334 overlaps between peaks and distal regulatory elements

#add distal targets to new peaks object
peaks_nd <- peaks_nearer
peaks_nd$target <- NULL
mcols(peaks_nd)[from(overlap_near),"target"] <- interactions[to(overlap_near)]$target
peaks_nd <- peaks_nd[!is.na(peaks_nd$target)]
mcols(peaks_nd)$target <- CharacterList(mcols(peaks_nd)$target)
head(peaks_nd)

#check result
#peaks_nearer[3] #seqname: 17 ranges: 903869-904158
#table(from(overlap_near)=="3") #the third peak should overlap with 7 targets
#table(from(overlap_near)=="1") #the first peak should overlap with 0 targets
#interactions[789097]$target #MYO1C
```

```{r}
#calculate percentage of peaks whose predicted regulatory targets are also the nearest TSS
#peaks_nd$nearestTSS.gene_name
#peaks_nd$target

#calculate how many peaks are equal and how many are not
table(any(peaks_nd$nearestTSS.gene_name == peaks_nd$target)) #48 peaks show the same distal target as the nearest TSS

#prepare new object with peaks that have distal regions
peaks_nd_distal <- peaks_nearer[unique(from(overlap_near))]  
#peaks_nd_distal #total 696 peaks

#prepare new object with peaks that show the same TSS
peaks_nd_distalTSS<- peaks_nd[which(any(peaks_nd$nearestTSS.gene_name == peaks_nd$target))]
#peaks_nd_distalTSS #total 48 peaks

#check specific peaks
#which(any(peaks_nd$nearestTSS.gene_name == peaks_nd$target))
#Example peak showing 
#peaks_nd[39]

#calculate proportion of all peaks
prop_near <- length(peaks_nd_distalTSS)/length(peaks_nd_distal)
#prop_near <- sum(any(peaks_nd$nearestTSS.gene_name == peaks_nd$target))/ sum(table(any(peaks_nd$nearestTSS.gene_name == peaks_nd$target))) 
prop_near #0.069
```


“Of the 696 peaks that are between 2.5 and 10kb from the nearest TSS and have a predicted target, 6.9 % form an interaction with that nearest gene. 

#### Calculate proportion of farer with predicted targets that equal the nearest TSS

```{r}
# find overlaps between the farer peaks and distal regulatory elements
overlap_far <- findOverlaps(peaks_farer, interactions)
#head(overlap_far)
overlap_far
length(overlap_far) #3336 overlaps between peaks and distal regulatory elements

#add distal targets to peaks object
peaks_fd <- peaks_farer
peaks_fd$target <- NULL
mcols(peaks_fd)[from(overlap_far),"target"] <- interactions[to(overlap_far)]$target
peaks_fd <- peaks_fd[!is.na(peaks_fd$target)]
mcols(peaks_fd)$target <- CharacterList(mcols(peaks_fd)$target)
head(peaks_fd)

#check result
#peaks_farer[19] 
#table(from(overlap_far)=="19") #the third peak should overlap with 1 target
#table(from(overlap_far)=="1") #the first peak should overlap with 0 targets
#interactions[802448]$target #HS3ST3A1
```

```{r}
#calculate percentage of peaks whose predicted regulatory targets are also the nearest TSS
#peaks_fd$nearestTSS.gene_name
#peaks_fd$target

#calculate how many peaks are equal and how many are not
table(any(peaks_fd$nearestTSS.gene_name == peaks_fd$target)) #85 peaks show the same distal target as the nearest TSS

#prepare new object with peaks that have distal regions
peaks_fd_distal <- peaks_farer[unique(from(overlap_far))]  
#peaks_fd_distal #total 1080 peaks

#prepare new object with peaks that show the same TSS
peaks_fd_distalTSS<- peaks_fd[which(any(peaks_fd$nearestTSS.gene_name == peaks_fd$target))]
#peaks_fd_distalTSS #total 84 peaks

#check specific peaks
#which(any(peaks_fd$nearestTSS.gene_name == peaks_fd$target))
#Example peak showing 
peaks_fd[44]

#calculate proportion of all peaks
prop_far <- length(peaks_fd_distalTSS)/length(peaks_fd_distal)
#prop_far <- sum(any(peaks_fd$nearestTSS.gene_name == peaks_fd$target))/ sum(table(any(peaks_fd$nearestTSS.gene_name == peaks_fd$target))) 
prop_far #0.078
```

“Of the 1080 peaks that are above 10kb from the nearest TSS and have a predicted target, 7.8 % form an interaction with that nearest gene. 

